Union and intersection contracts are hard, actually

### 2021-06-09 paper.tex
Rework section 5.3 [sec 5 finished]

Racket is a language based on Scheme, mainly thought as a language to experiment
with different language desings and ideas. It provides a very complete contract
system\cite{RacketContracts}.

Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
These contracts are lightweight versions of intersections and unions, they simply
check that every contract is valid, in the case of \racket{and/c},
and that at least one contract
is valid, in the case of \racket{or/c}.

So, for instance, the usual usecase of intersection to encode bounded overloading
does not work on Racket. Take a look at the next example, the function
\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
since 3 is not a valid \racket{string?}.

@@

\begin{lstlisting}[language=racket]
(define/contract overload
(and/c (-> number? number?)
       (-> string? string?))
(lambda (x) x))
\end{lstlisting}
\caption{Non working function overloading on Racket.}

Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
contracts that could (potentially) apply to a function, these have to be
differentiated by first order markers.
For instance, the code on Figure~\ref{code:racket:or/c:working}
would correctly allow only to call
\racket{united} with a \racket{number?},
and it can make this distinction since it knows that
\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
before running, since the \racket{or/c} contract is unable to know
which of the two function contracts to use.
Ideally, this last example should just behave as wrapping
the function with the \racket{(-> even? even?)} contract.

\begin{lstlisting}[language=racket]
(define/contract united
(or/c (-> number? number?)
      (-> string? string? string?))
(lambda (x) x))
\end{lstlisting}
\caption{Working \racket{or/c} example on Racket.}
\label{code:racket:or/c:working}

@@

\begin{lstlisting}[language=lisp]
(define/contract united
(or/c (-> number? number?)
      (-> even? even?))
(lambda (x) x))
\end{lstlisting}
\caption{Non working \racket{or/c} example on Racket.}
\label{code:racket:or/c:non-working}

@@

Racket does provide an alternative for overloading methods, called the
\racket{case->} contract, however, this overloading of methods only
works over contracts of functions with different amount of parameters.
It's a good alternative to allow limited overloading of functions.
As an example, consider the following non failing piece of code, and
try to understand how Racket can decide, at each application,
which function contract to use.

\begin{lstlisting}[language=racket]
(define/contract overcase
  (case-> (-> string? string?)
          (-> number? number? number?)
          )
  (lambda (x [y 0]) (if (number? x)
                   (+ x y)
                   x)))

@@

(overcase "hello")
\end{lstlisting}
\end{figure}

Even if Racket's intent is not to implement union and intersection
dynamic checks, they do provide contracts that are, in some sense,
simplified versions of those.
In our eyes, Racket has accepted the difficulty of correctly and practically
combine union, intersection, and higher order contracts, and has choose
to prioritise the highe order ones.

>>>>

Racket is a language based on the Scheme dialect of Lisp. Among established
languages, Racket is probably the one with the most comprehensive contract
system\cite{RacketContracts}. Regarding union and intersections, Racket provides
the \racket{and/c} and \racket{or/c} combinators for contract.  These
combinators correspond to the naive interpretation of unions and intersections
(as in Figure~\ref{fig:naive-impl}): \racket{and/c} (resp.  \racket{or/c})
simply checks sequentially that all (resp. one) contracts pass.  In consequence,
unions or intersections of higher-order contracts don't match intuition, and in
particular intersections don't model overloading. As in
Figure~\ref{fig:naive-impl}, the example given in
Figure~\ref{fig:racket-overloading} always fail because no argument satisfies
both \racket{number?} and \racket{string?}.

% Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
% These contracts are lightweight versions of intersections and unions, they simply
% check that every contract is valid, in the case of \racket{and/c},
% and that at least one contract
% is valid, in the case of \racket{or/c}.

% So, for instance, the usual usecase of intersection to encode bounded overloading
% does not work on Racket. Take a look at the next example, the function
% \racket{overload} should accept a call like \racket{(overload 3)}, but it fails
% since 3 is not a valid \racket{string?}.

\begin{figure}[h]
\begin{lstlisting}[language=racket]
(define/contract overload
(and/c (-> number? number?)
       (-> string? string?))
(lambda (x) x))
\end{lstlisting}
\caption{\racket{and/c} and overloading}
\label{fig:racket-overloading}
\end{figure}

When higher-order contracts are combined using \racket{or/c}, Racket imposes
that contracts must distinguishable by their arity. Doing so, there is at most
one candidate that can be selected directly. This is illustrated in
Figure~\ref{code:racket:or/c:working} whose program is accepted. On the other
hand, the program of Figure~\ref{code:racket:or/c:non-working} is rejected.

% that could (potentially) apply to a function, these have to be
% differentiated by first order markers.
% For instance, the code on Figure~\ref{code:racket:or/c:working}
% would correctly allow only to call
% \racket{united} with a \racket{number?},
% and it can make this distinction since it knows that
% \racket{(lambda (x) x)} can not be a function that takes 2 parameters.
% However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
% before running, since the \racket{or/c} contract is unable to know
% which of the two function contracts to use.
% Ideally, this last example should just behave as wrapping
% the function with the \racket{(-> even? even?)} contract.

@@

\begin{lstlisting}[language=racket]
(define/contract united
(or/c (-> number? number?)
      (-> string? string? string?))
(lambda (x) x))
\end{lstlisting}
\caption{Accepted use of \racket{or/c} with higher-order contracts}
\label{code:racket:or/c:working}

@@

\begin{lstlisting}[language=lisp]
(define/contract united
(or/c (-> number? number?)
      (-> even? even?))
(lambda (x) x))
\end{lstlisting}
\caption{Rejected use of \racket{or/c} with higher-order contracts}
\label{code:racket:or/c:non-working}

@@

Racket does provide the \racket{case->} combinator which(partially) supports
overloading. As for the \racket{or/c}, the candidate contracts must have
distinct arities to avoid ambiguity. An example is provided in
Figure~\ref{fig:rafig:racket-case-fun}. The resulting possibilities are similar
to static overloading, where one function can take additional parameters for
example (e.g. as supported for Java methods). Unfortunately, it excludes the
overloading of generic operations with fixed arity such as equality, comparison,
arithmetic operators, and so on.

% Racket does provide an alternative for overloading methods, called the
% \racket{case->} contract, however, this overloading of methods only
% works over contracts of functions with different amount of parameters.
% It's a good alternative to allow limited overloading of functions.
% As an example, consider the following non failing piece of code, and
% try to understand how Racket can decide, at each application,
% which function contract to use.

\begin{figure}[h]
\begin{lstlisting}[language=racket]
(define/contract overcase
  (case-> (-> string? string?)
          (-> number? number? number?)
          )
  (lambda (x [y 0]) (if (number? x)
                   (+ x y)
                   x)))

@@

(overcase "hello")
\end{lstlisting}
\caption{Overloading with the \racket{case->} combinator}
\label{fig:racket-case-fun}
\end{figure}

In conclusion, Racket also seems to avoid the difficulties of general unions and
intersections. They make the pragmatic choice of a simple semantics that work
for non higher-order contracts and a limited support for overloading.
% Even if Racket's intent is not to implement union and intersection
% dynamic checks, they do provide contracts that are, in some sense,
% simplified versions of those.
% In our eyes, Racket has accepted the difficulty of correctly and practically
% combine union, intersection, and higher order contracts, and has choose
% to prioritise the highe order ones.
